---
title: "MR-rr workflow for multiple exposures and outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MR-rr workflow for multiple exposures and outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Overview

`MR.rr` estimates a low-rank causal effect matrix from summary-level genetic
association data with multiple exposures and multiple outcomes. The package
contains four estimators:

- `"naive"`: reduced-rank estimation without measurement-error correction;
- `"corrected"`: the main measurement-error-corrected MR-rr estimator;
- `"regularized"`: a spectrally regularized MR-rr estimator; and
- `"sparse"`: sparse MR-rr support selection with an optional unpenalized
  post-selection refit.

This vignette describes the user-facing analysis workflow. Manuscript-specific
simulation, bootstrap, cluster-submission, and figure-generation scripts are
not part of the package interface.

```{r load-package}
library(MR.rr)
```

## Input structure

Suppose the analysis contains `m` genetic instruments, `p` exposures, and
`q` outcomes. The four required matrices are:

| Argument | Dimension | Contents |
|---|---:|---|
| `beta_exposure` | `m` by `p` | SNP-exposure association estimates |
| `se_exposure` | `m` by `p` | SNP-exposure standard errors |
| `beta_outcome` | `m` by `q` | SNP-outcome association estimates |
| `se_outcome` | `m` by `q` | SNP-outcome standard errors |

Every row must refer to the same genetic instrument in the same order across
all four matrices. The estimates must already be harmonized to the same effect
allele. `MR.rr` does not perform allele harmonization, LD clumping, or
instrument selection.

Optional inputs include:

- `cor_exposure`, a `p` by `p` exposure-trait correlation matrix;
- `cor_outcome`, a `q` by `q` outcome-trait correlation matrix;
- `allele_frequency`, one effect-allele or minor-allele frequency per
  instrument;
- `variant_variance`, one genotype variance per instrument, supplied instead
  of `allele_frequency`; and
- `W`, a positive-definite `q` by `q` outcome weight matrix.

When allele frequencies are supplied, the genotype variance is calculated as

\[
2 f_j(1-f_j).
\]

When `W = NULL`, the inverse of the constructed outcome measurement-error
covariance matrix is used.

## Construct an example dataset

The following example has a rank-one causal effect matrix and eight exposures,
of which three have nonzero coefficients in the latent pathway.

```{r simulate-data}
set.seed(20260825)

m <- 300L
p <- 8L
q <- 3L

instrument_names <- paste0("rs", seq_len(m))
exposure_names <- paste0("exposure_", seq_len(p))
outcome_names <- paste0("outcome_", seq_len(q))

beta_exposure <- matrix(
  rnorm(m * p, sd = 0.10),
  nrow = m,
  ncol = p,
  dimnames = list(instrument_names, exposure_names)
)

true_outcome_loading <- c(0.80, -0.50, 0.30)
true_exposure_coefficients <- c(0.30, -0.20, 0, 0.15, 0, 0, 0, 0)

true_effect <- outer(
  true_outcome_loading,
  true_exposure_coefficients
)
dimnames(true_effect) <- list(outcome_names, exposure_names)

beta_outcome <- beta_exposure %*% t(true_effect) +
  matrix(rnorm(m * q, sd = 0.01), nrow = m, ncol = q)
dimnames(beta_outcome) <- list(instrument_names, outcome_names)

se_exposure <- matrix(
  0.01,
  nrow = m,
  ncol = p,
  dimnames = list(instrument_names, exposure_names)
)

se_outcome <- matrix(
  0.015,
  nrow = m,
  ncol = q,
  dimnames = list(instrument_names, outcome_names)
)

allele_frequency <- setNames(
  runif(m, min = 0.10, max = 0.50),
  instrument_names
)
```

## Prepare the data

`prepare_mr_rr_data()` validates dimensions and names, applies optional
genotype-variance scaling, constructs the average exposure and outcome
measurement-error covariance matrices, and returns an `mr_rr_data` object.

```{r prepare-data}
prepared <- prepare_mr_rr_data(
  beta_exposure = beta_exposure,
  se_exposure = se_exposure,
  beta_outcome = beta_outcome,
  se_outcome = se_outcome,
  allele_frequency = allele_frequency
)

c(
  instruments = prepared$n_instruments,
  exposures = prepared$n_exposures,
  outcomes = prepared$n_outcomes
)
```

The matrices used by the estimators are available as `prepared$X`,
`prepared$Y`, `prepared$Sigma_X`, and `prepared$W`.

## Fit several estimators

`fit_all_mr_rr()` applies several estimators to the same prepared data using a
common working rank.

```{r fit-estimators}
fits <- fit_all_mr_rr(
  data = prepared,
  methods = c("naive", "corrected", "regularized"),
  rank = 1,
  regularization_rate = 1e-6
)

fits
round(fits$corrected$AB, digits = 3)
```

Each fitted estimator contains:

- `A`, the outcome loading matrix;
- `B`, the exposure coefficient matrix; and
- `AB`, the estimated causal effect matrix.

The method-specific results are accessed by name, such as
`fits$naive$AB`, `fits$corrected$AB`, and `fits$regularized$AB`.

## Select the working rank

Set `rank = NULL` to apply the sequential rank diagnostic. In a collection of
fits, rank selection is performed once and the selected rank is reused by all
requested estimators.

```{r rank-selection}
fits_selected <- fit_all_mr_rr(
  data = prepared,
  methods = c("naive", "corrected", "regularized"),
  rank = NULL,
  rank_alpha = 0.05,
  regularization_rate = 1e-6
)

fits_selected$rank

rank_diagnostics <- with(
  fits_selected$rank_selection,
  data.frame(
    candidate_rank = candidate_ranks,
    statistic = round(statistic, digits = 3),
    df = df,
    p_value = signif(p_value, digits = 3)
  )
)

rank_diagnostics
```

This procedure is a working rank-selection diagnostic. Its chi-square
approximation should be interpreted cautiously when instruments are weak.

## Fit sparse MR-rr

Sparse MR-rr uses an L1 penalty to select the support of `B`. With
`sparse_refit = TRUE`, the selected support is then held fixed while the model
is refitted without the L1 penalty. The refit reduces shrinkage bias without
changing the selected support.

The sparse example is not evaluated while building the vignette because it
requires a convex optimization solver. It can be run interactively after
installing `CVXR`.

```{r sparse-estimator, eval = FALSE}
sparse_fit <- fit_mr_rr(
  data = prepared,
  method = "sparse",
  rank = 1,
  sparse_lambda = 1e-2,
  sparse_refit = TRUE,
  sparse_threshold = 1e-2
)

sparse_comparison <- data.frame(
  exposure = exposure_names,
  true_support = true_exposure_coefficients != 0,
  selected_support = as.vector(sparse_fit$sparse_selection$B != 0),
  penalized_B = round(
    as.vector(sparse_fit$sparse_selection$B),
    digits = 3
  ),
  refitted_B = round(as.vector(sparse_fit$B), digits = 3)
)

sparse_comparison
round(sparse_fit$AB, digits = 3)
```

The relevant sparse components are:

- `sparse_fit$sparse_selection$B_raw`: penalized coefficients before the final
  numerical threshold;
- `sparse_fit$sparse_selection$B`: thresholded penalized coefficients defining
  the selected support;
- `sparse_fit$B`: coefficients after the optional unpenalized refit; and
- `sparse_fit$AB`: the final estimated causal effect matrix.

## Interpret the factorization

The causal effect matrix is represented as

\[
C = AB.
\]

The scales of `A` and `B` depend on the normalization imposed on the
factorization. Consequently, a numerically large entry of `B` does not by
itself imply a large causal effect. The primary causal estimand is `AB`.
`B` is additionally useful for interpreting latent exposure pathways and, in
the sparse estimator, the selected support.

## Regularized implementation

The default regularized estimator evaluates the spectral filter directly and
avoids explicitly inverting the unstable corrected exposure matrix:

```{r regularized-estimator}
regularized_fit <- fit_mr_rr(
  data = prepared,
  method = "regularized",
  rank = 1,
  regularization_rate = 1e-6,
  regularized_implementation = "spectral"
)

round(regularized_fit$AB, digits = 3)
```

The historical implementation remains available through
`regularized_implementation = "legacy"` for controlled comparisons. The
spectral implementation is the package default because it avoids two unstable
matrix inversions.

## Manuscript reproducibility

The package provides the maintained user-facing estimators. Exact manuscript
replication uses the frozen analysis scripts, data, results, checksums, and
recorded software environment under
`freeze/current_analysis_20260823/project/`. The analysis freeze must remain
unchanged even when the package implementation is improved.
